---
title: "Multivariate Analysis - PCA and EFA"
author: "Alexis Alberto Zuniga Alonso"
date: today
format:
  html:
    theme: flatly
    toc: true
    code-fold: show
    code-tools: true
    embed-resources: true
    fig-width: 8
  typst:
    toc: true
    fig-width: 7
execute:
  warning: true
  message: true
  echo: true
---

# 1. Introduction
Analyse the variables below from the dataset 'Holzinger-Swineford 1939' with multivariate methods such as Principal Component Analysis (PCA for short) and Exploratory Factor Analysis (EFA).

1. Visual (x1)
2. Cubos (x2)
3. Índice de Lozenges (x3)
4. Compresión de párrafos (x4)
5. Completar oraciones (x5)
6. Significado de palabras (x6)
7. Velocidad para sumar (x7)
8. Velocidad para contar puntos (x8)
9. Discriminación aceleradad de mayúsculas rectas y curvas (x9)

## 1.1. Project Libraries and Installation

```{r}
libs <- c("lavaan", "psych", "GPArotation", "FactoMineR",
  "factoextra", "corrplot", "skimr", "tidyverse")

to_install <- setdiff(libs, rownames(installed.packages()))
if (length(to_install)) install.packages(to_install)

invisible(lapply(libs, library, character.only = TRUE))
```

Then, we import the library `lavaan` where the dataset already is.

```{r}
library(lavaan)

data <- HolzingerSwineford1939
```


We first glimpse at the data with `glimpse(data)` and `skim(data)`

```{r}
glimpse(data)
skim(data)
```

---

We also display a correlation matrix using the function `corrplot(cor(df), method = "number")` that shows how correlated our variables `x1` to `x9` are.
```{r}
df <- data |> select(x1:x9)
corrplot(cor(df), method = "number")
```

